Measuring contrast processing in the visual system using SSVEP (and VEP?)

### 2024-04-03 review.qmd
Changed citation style

---
title: "Measuring contrast processing in the visual system using SSVEP (and VEP?)"
format: pdf
bibliography: WadeBaker_Review.bib
execute:
  echo: false
---

@@

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for at least 100 years. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years[@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This document can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

## INTRODUCTION:
Neurons in the visual areas of the brain are primarily responsive to changes in cone photoreceptor activations across time and space. This property, referred to as contrast, sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity [@Campbell1965], defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di Russo, Hillebrand:]. The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. SSVEP signals can also be elicited by periodic changes of stimulus properties other than achromatic and chromatic contrast, such as, motion, stereo depth, and facial identity or expression [see @Norcia2015, for an overview]; however our focus here is on the contrast response.

### Why measure responses to contrast?
Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over time ('spatial contrast') or space ('spatial contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well. [McKeefry, Sutter, Baseler, DiRusso...].

Contrast is relatively simple to define: typically, contrast is specified as the percentage deviation of uniform stimulus from the background. So, for example, a disk of 100 units, $I_{\mathrm{stim}}$ of cone activation surrounded by a 'background' of 50 units of activation $I_{\mathrm{background}}$ has a contrast of $\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}$ = 100%. Where patterns are more complex (for example, the sine-wave gratings or Gabor patches common in vision science), the Michaelson definition of contrast is specified by the maximum and minimum excursions from the mean: $${{\frac {I_{\mathrm {stimmax} }-I_{\mathrm {stimmin} }}{I_{\mathrm {stimmax} }+I_{\mathrm {stimmin} }}},} $$. These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance [LenniePokornySmith]) and also to definitions based on cone excitations [DKL, MacBoynton] which are more common in work on chromatic processing.

Although its definition as a stimulus is simple, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [Nachmas and Sansbury? Foley? Heeger 92, Heeger and Carandini 2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background.
```{python}

#| label: Perceptual gain control
#| fig-cap: The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region"
#| fig-cap-location: bottom
#|
import numpy as np
import matplotlib.pyplot as plt

@@

 Individual neurons at multiple stages of the visual hierarchy also change their sensitivity depending on the average spatiotemporal contrast energy of their environment. This "normalisation" process is dynamic and nonlinear and is well-modeled by a hyperbolic ratio function in which the response of each neuron is modulated by a local 'gain pool' composed of the summed respnses of the local neuronal population.[@Heeger 92, @Carandini and Heeger 2011 @Busse 2008, @Baker and Wade] 
Individual neurons at multiple stages of the visual hierarchy also change their sensitivity depending on the average spatiotemporal contrast energy of their environment. This "normalisation" process is dynamic and nonlinear and is well-modeled by a hyperbolic ratio function in which the response of each neuron is modulated by a local 'gain pool' composed of the summed responses of the local neuronal population [@Heeger1992; @CarandiniHeeger2011; @Busse2008; @BakerWade].

@@

Here, we show the effect of adding a second contrast component: in general, the two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms --- is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan].
First, we implement a simple non-linear transducer with a single input to illustrate the typical signoidal shape of a contrast response function.
(Figure xx)

@@

#| label: plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain"
#| fig-cap-location: bottom
#| 
import matplotlib.pyplot as plt
import numpy as np
import scipy as scp

@@

# Adjust the spacing between subplots
plt.tight_layout()
plt.show()
```

@@

#| label: plotspectra
#| fig-cap: "Temporal frequency representations of combining two inputs at frequencies F1=5Hz and F2=7Hz combination. A pure sine-wave input would generate responses only at F1 and F2. However, the rectification that accompanies early visual processing converts this contrast reversal into a more complex frequency-domain representation with power predominantly at the second harmonics 2F1 and 2F2"
#| fig-cap-location: bottom

#| label: plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain"
#| fig-cap-location: bottom
#| 
import matplotlib.pyplot as plt
import numpy as np
import scipy as scp

@@

#| label: ContrastResponseFunction
#| fig-cap: "Non-linear contrast response function resulting from the population average of two full-wave rectified signals passing through a non-linear transducer. Note the 'supersaturation' at high contrast levels."
#| fig-cap-location: bottom

@@

This possibility was recognised in the early days of the technique [@reganandRegan1988]. Although characterising the complete set of computations along entire processing pathway is problematic, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. Candy and Norcia (in infants) 2001, Busse et al2008, Tsai/Norcia/Wade 2010s, Then lots of Baker stuff including the CerebCort paper. : Focusing only on SSVEP: [@NorciaCandy2001, @Regan1990?  ]

@@

The sweep VEP (technically, the sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. Because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004 Diva vs Enfant, @Hoffman, @Bach and @Farmer].
The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffman; @Bach; @Farmer].

@@

>>>>

---
title: "Measuring contrast processing in the visual system using SSVEP (and VEP?)"
format: pdf
bibliography: WadeBaker_Review.bib
csl: visual-neuroscience.csl
execute:
  echo: false
---

@@

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for at least 100 years. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR, and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This document can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

## INTRODUCTION

Neurons in the visual areas of the brain are primarily responsive to changes in cone photoreceptor activations across time and space. This property, referred to as contrast, sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity [@Campbell1965], defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2007] (Hillebrand?). The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. SSVEP signals can also be elicited by periodic changes of stimulus properties other than achromatic and chromatic contrast, such as motion, stereo depth, and facial identity or expression [see @Norcia2015, for an overview]; however our focus here is on the contrast response.

### Why measure responses to contrast?

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well. [McKeefry, Sutter, Baseler, DiRusso...].

Contrast is relatively simple to define: typically, contrast is specified as the percentage deviation of a uniform stimulus from the background. So, for example, a disk of 100 units of cone activation ($I_{\mathrm{stim}}$) surrounded by a 'background' of 50 units of activation ($I_{\mathrm{background}}$) has a contrast of $\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}$ = 100%. Where patterns are more complex (for example, the sine-wave gratings or Gabor patches common in vision science), the @Michelson1927 definition of contrast is specified by the maximum and minimum excursions from the mean: $${{\frac {I_{\mathrm {stimmax} }-I_{\mathrm {stimmin} }}{I_{\mathrm {stimmax} }+I_{\mathrm {stimmin} }}}.} $$ These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance [LenniePokornySmith]) and also to definitions based on cone excitations [@Derrington1984, @MacLeod1979] which are more common in work on chromatic processing.

Although its definition as a stimulus is simple, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background.

```{python}
#| label: Perceptual gain control
#| fig-cap: The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region"
#| fig-cap-location: bottom

import numpy as np
import matplotlib.pyplot as plt

@@

Individual neurons at multiple stages of the visual hierarchy also change their sensitivity depending on the average spatiotemporal contrast energy of their environment. This "normalisation" process is dynamic and nonlinear and is well-modeled by a hyperbolic ratio function in which the response of each neuron is modulated by a local 'gain pool' composed of the summed respnses of the local neuronal population.[@Heeger 92, @Carandini and Heeger 2011 @Busse 2008, @Baker and Wade]

Individual neurons at multiple stages of the visual hierarchy also change their sensitivity depending on the average spatiotemporal contrast energy of their environment. This "normalisation" process is dynamic and nonlinear and is well-modeled by a hyperbolic ratio function in which the response of each neuron is modulated by a local 'gain pool' composed of the summed responses of the local neuronal population [@Heeger1992; @CarandiniHeeger2011; @Busse2008; @BakerWade].

@@

Here, we show the effect of adding a second contrast component: in general, the two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms --- is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan].

First, we implement a simple non-linear transducer with a single input to illustrate the typical signoidal shape of a contrast response function.
(Figure xx)

@@

```{python}
#| label: plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain"
#| fig-cap-location: bottom

import matplotlib.pyplot as plt
import numpy as np
import scipy as scp

@@

# Adjust the spacing between subplots
plt.tight_layout()
plt.show()

@@

```{python}
#| label: plotspectra
#| fig-cap: "Temporal frequency representations of combining two inputs at frequencies F1=5Hz and F2=7Hz combination. A pure sine-wave input would generate responses only at F1 and F2. However, the rectification that accompanies early visual processing converts this contrast reversal into a more complex frequency-domain representation with power predominantly at the second harmonics 2F1 and 2F2"
#| fig-cap-location: bottom

#| label: plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain"
#| fig-cap-location: bottom

import matplotlib.pyplot as plt
import numpy as np
import scipy as scp

@@

```{python}
#| label: ContrastResponseFunction
#| fig-cap: "Non-linear contrast response function resulting from the population average of two full-wave rectified signals passing through a non-linear transducer. Note the 'supersaturation' at high contrast levels."
#| fig-cap-location: bottom

@@

This possibility was recognised in the early days of the technique [@reganandRegan1988]. Although characterising the complete set of computations along entire processing pathway is problematic, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. Candy and Norcia (in infants) 2001, Busse et al2008, Tsai/Norcia/Wade 2010s, Then lots of Baker stuff including the CerebCort paper. : Focusing only on SSVEP: [@NorciaCandy2001, @Regan1990]

@@

The sweep VEP (technically, the sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. Because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004 Diva vs Enfant, @Hoffman, @Bach and @Farmer].

The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffman; @Bach; @Farmer].

@@

# References
